\pagestyle{fancy}
# COVID outcome prediction in the emergency department (COPE) {#sec-covid}

\blfootnote{Chapter based on \textit{van Klaveren D, Rekkas A, Alsma J, et al. COVID outcome prediction in the emergency department (COPE): using retrospective Dutch hospital data to develop simple and valid models for predicting mortality and need for intensive care unit admission in patients who present at the emergency department with suspected COVID-19BMJ Open 2021;11:e051468. doi: 10.1136/bmjopen-2021-051468}}

\chaptermark{COVID Outcome Prediction in the emergency department}
\vspace*{\fill}\par
\pagebreak

## Abstract {.unnumbered}

**Objectives**: Develop simple and valid models for predicting mortality and need for
intensive care unit (ICU) admission in patients who present at the emergency
department (ED) with suspected COVID-19. 

**Design**: Retrospective. 

**Setting**: Secondary care in four large Dutch hospitals.

**Participants**:
Patients who presented at the ED and were admitted to hospital with suspected
COVID-19. We used 5831 first-wave patients who presented between March and
August 2020 for model development and 3252 second-wave patients who presented
between September and December 2020 for model validation. Outcome measures We
developed separate logistic regression models for in-hospital death and for need
for ICU admission, both within 28 days after hospital admission. Based on prior
literature, we considered quickly and objectively obtainable patient
characteristics, vital parameters and blood test values as predictors. We
assessed model performance by the area under the receiver operating
characteristic curve (AUC) and by calibration plots.

**Results**:
Of 5831 first-wave patients, 629 (10.8%) died within 28 days after admission.
ICU admission was fully recorded for 2633 first-wave patients in 2 hospitals,
with 214 (8.1%) ICU admissions within 28 days. A simple model with age,
respiratory rate, C reactive protein, lactate dehydrogenase, albumin and urea
captured most of the ability to predict death. COPE was well calibrated and
showed good discrimination for mortality in second-wave patients (AUC in four
hospitals: 0.82 (95% CI 0.78 to 0.86); 0.82 (95% CI 0.74 to 0.90); 0.79 (95% CI
0.70 to 0.88); 0.83 (95% CI 0.79 to 0.86)). COPE was also able to identify
patients at high risk of needing ICU admission in second-wave patients (AUC in
two hospitals: 0.84 (95% CI 0.78 to 0.90); 0.81 (95% CI 0.66 to 0.95)).

**Conclusions**:
COPE is a simple tool that is well able to predict mortality and need for ICU
admission in patients who present to the ED with suspected COVID-19 and may help
patients and doctors in decision making.

\vspace*{\fill}\par
\pagebreak

## Background {.unnumbered}

The COVID-19 pandemic is putting extraordinary pressure on emergency departments
(EDs), clinical wards and intensive care units (ICUs). Clinical prediction
models for COVID-19 outcomes have the potential to support decision making about
hospital admission. Existing models that predict mortality for non-trauma
patients presenting to the ED are unlikely to be well calibrated and optimally
discriminating for patients with COVID19 [@brink2019prediction]. Most currently
available models specifically developed for patients with COVID-19 that were
assessed with the prediction model risk of bias assessment tool contain a high
risk of bias [@wynants2020prediction;@sperrin2020prediction;@moons2019probast]. The most common reasons were non-representative selection of
control patients, exclusion of patients in whom the event of interest was not
observed by the end of the study, high risk of model overfitting and vague
reporting. Additionally, the description of the study population or intended use
of the models was often missing, and calibration of the model predictions was
rarely assessed.

The recently proposed 4C Mortality Score is probably at low risk of bias, but
was derived from a selected population of patients admitted to UK hospitals who
were seriously ill (mortality rate of 32.2%). Predictors included the number of
comorbidities and the Glasgow Coma Scale, items that are not easily and
unambiguously obtained for patients with suspected COVID-19 at EDs everywhere
[@knight2020risk;@docherty2020features]. Similarly, the promising risk scores
Veterans Health Administration COVID-19 (VACO) and COVID-GRAM—predicting 30-day
mortality in positively tested patients and critical illness in hospitalised
patients, respectively—require knowledge on pre-existing comorbidities
[@liang2020development;@king2020development]. The COVID-GRAM model also requires
chest radiography results.

We aimed to develop and validate a simple and valid model for predicting
mortality and the need for ICU in all patients who are suspected to have
COVID-19 when presenting at the ED. To facilitate implementation in clinical
practice, we only included quickly and objectively obtainable patient
characteristics, vital parameters and blood test values.

## Methods {.unnumbered}

### Population {.unnumbered}

Nineteen large Dutch hospitals were requested to supply anonymised retrospective
data on the cohorts of patients with COVID-19 who were admitted to their
hospital. Of those hospitals, Catharina Hospital Eindhoven, Zuyderland Medical
Center Heerlen, Isala Clinics Zwolle, Erasmus University Medical Center
Rotterdam and Antonius Hospital Sneek supplied these data. The data from
Antonius Hospital Sneek were not used in the analyses, because of large
proportions of missing predictor values.

For model development, we used the data of patients who presented at the ED and
were admitted to the hospital with suspected COVID-19 in the first wave of the
pandemic, that is, from March up to and including August 2020. Patients being
transferred to other hospitals were excluded since information on outcomes was
missing. For model validation, we used data of patients who presented at the ED
and were admitted to the hospital with suspected COVID-19 in the second wave of
the pandemic, that is, from September up to and including December 2020.
Potential multiple hospital admissions of the same patient were considered as
independent hospital admissions.

### Outcomes {.unnumbered}

The outcomes of interest were: (1) in-hospital death or transfer to a hospice
within 28 days after hospital admission and (2) admission to ICU within 28 days
after hospital admission.

### Predictors {.unnumbered}

Based on prior literature, we included patient characteristics (sex, age, body
mass index), vital parameters (oxygen saturation, systolic blood pressure, heart
rate (HR), respiratory rate (RR), body temperature) and blood test values (C
reactive protein (CRP), lactic dehydrogenase (LDH), D-Dimer, leucocytes,
lymphocytes, monocytes, neutrophils, eosinophils, Mean Corpuscular Volume (MCV),
albumin, bicarbonate, sodium, creatinine, urea), all measured at ED admission,
as potential predictors [@wynants2020prediction]. Furthermore, we included the
month of admission to capture potential changes in outcomes over time. In case
of multiple measurements for the same patient, we used the first measurement
after presentation at the ED. We used multivariate imputation by chained
equations (R-packages mice) for multiple imputation of missing predictor values
[@r_core_team2020project;@van_buuren2011mice]. Multiple imputation in the
validation data was undertaken separately from multiple imputation in the
development data to ensure fully independent model validation.

### Model development {.unnumbered}

Logistic regression was used to analyse associations between predictors and
outcomes. We decided on including non-linear transformations of potential
predictors on the basis of a full model with a restricted cubic spline (three
knots; two regression coefficients) for each continuous
predictor[@stone1985additive;@harrell2015regression]. Based on Wald statistics,
we selected the most promising predictors into a parsimonious model for easy use
in clinical practice. To prevent overfitting, we used bootstrap
validation—including the same variable selection strategy to mimic our modelling
strategy—to estimate a uniform shrinkage factor [@harrell2015regression]. The
regression coefficients of the final model were multiplied by this shrinkage
factor, and the model intercept was adjusted to ensure overall calibration of
the model. We used the R-package rms (Regression Modelling Strategies) for
regression analyses [@r_core_team2020project;@jr2023rms].

### Model validation {.unnumbered}

Model performance was assessed with temporal validation in second wave patients,
in each of the four separate hospitals. We assessed discriminative ability with
the area under the receiver operating characteristic curve (AUC) and calibration
with calibration plots of five equally sized groups of predicted risk,
calibration intercepts and calibration slopes. The model-based concordance (mbc)
was used to understand the impact of potential differences in case-mix
heterogeneity between the development and validation data on discriminative
ability [@van_klaveren2016new].

### Patient and public involvement {.unnumbered}

Patients were not directly involved in the design of this study. The outcome of
interest and the potential predictors were selected up front by a group of
hospital physicians caring for patients with COVID-19 (ED, internal medicine,
pulmonary medicine, ICU). Since we retrospectively collected data, patients were
not burdened by our study. In future research, we will convene multi-stakeholder
panels of approximately 12 members including patients with COVID-19, relatives,
hospitals physicians caring for patients with COVID-19, palliative care
physicians and ethicists, with the aim to develop a full understanding of how
the models may best support patients and clinicians in making critical
patient-centred decisions.

## Results {.unnumbered}
### Population and outcomes {.unnumbered}

The database contained 5912 patients who presented at the ED from March up to
and including August 2020 and who were admitted to the hospital with a suspicion
of COVID-19. Of those patients 81 (1.4%) were excluded because of a transfer to
other hospitals (outcome not recorded). The development data included 5831
patients of whom 629 (10.8%) died, 5070 (86.9%) were discharged within 28 days
after hospital admission, and 132 (2.3%) were still in hospital at 28 days after
admission. Patients who died—in comparison with patients who were
discharged—tended to be more often male (64% vs 56%), at older age (median 78 vs
69), with higher RR (median 23 vs 19) and HR (median 93 vs 90), lower oxygen
saturation (median 94.1 vs 96.0), higher blood levels of CRP (median 91 vs 43),
LDH (median 338 vs 237), creatinine (median 102 vs 82) and urea (median 9.6 vs
6.2) and lower blood levels of lymphocytes (median 0.80 vs 1.10) and albumin
(median 36 vs 40) (Table \ref{table:ch6-table1}). Similar patterns were seen in
3252 patients who were admitted to hospital in the second wave of the pandemic
from September up to and including December 2020, of whom 326 (10.0%) died, 2854
(87.8%) were discharged within 28 days after admission, and 72 (2.2%) were still
in hospital at 28 days after admission.

Admission to ICU was fully recorded—including ICU admissions at a later time
point than the initial hospital admission—for 2633 patients in 2 hospitals (214
ICU admissions within 28 days (8.1%)) in the first wave of the pandemic.
Patients who were admitted to the ICU—in comparison with patients who were
discharged or died without being admitted to the ICU—tended to be more often
male (68% vs 57%), with higher RR (median 23 vs 19) and HR (median 91 vs 88),
lower oxygen saturation (median 95.0 vs 95.8), higher blood levels of CRP
(median 88 vs 47), LDH (median 318 vs 234), creatinine (median 93 vs 84) and
urea (median 7.1 vs 6.6) and lower blood levels of albumin (median 38 vs 40)
(Table \ref{table:ch6-table2}). In contrast with patients who died, patients who
were admitted to the ICU were not older than patients who were discharged
(median 68 vs 71), probably due to decisions not to admit frail patients to the
ICU. Patterns were similar in 1466 patients (86 ICU admissions within 28 days
(5.9%)) who were admitted to these 2 hospitals in the second wave of the
pandemic.

\begin{landscape}
\begin{table}[h]
\caption{Baseline characteristics of development and validation patient cohorts
median (‘M’) and quartile range (‘Q1’=first quartile; ‘Q3’=third quartile) are
presented for all continuous variables (continues to next page).}
\label{table:ch6-table1}
\footnotesize
\resizebox{\columnwidth}{!}{\begin{tabular}{lrrrrrrrrrrrrrr}
\toprule
& \% missing & All patients &   Discharged or dead &  In hospital & ICU admission \\
\midrule
\textbf{A. Development data} &  & n = 5831 & n = 5070 & n = 132 & n = 629 \\
Male sex & 0 & 57\%   & 56\% & 64\% & 64\%  \\
Age (years) & 0 & 70 (58, 80) & 69 (56, 78) & 71 (62, 79) & 78 (70, 84) \\
BMI (kg/m²) & 58 & 26 (23, 30) & 26 (23, 30) & 25 (23, 29) & 26 (23, 30) \\
HR (bpm) & 39 & 90 (78, 103) & 90 (78, 103) & 87 (76, 99) & 93 (80, 107) \\
SBP (mmHg) & 42 & 133 (118, 150) & 133 (119, 151) & 136 (115, 152) & 131 (114, 145) \\
RR (/min) & 42 & 19 (16, 23) & 19 (16, 23) & 20 (17, 24) & 23 (19, 28) \\
Saturation (\%) & 41 & 95.8 (94, 97.5) & 96 (94.3, 97.8) & 95.4 (93.6, 97) & 94.1 (91.9, 96) \\
Temperature (°C) & 40 & 37.3 (36.7, 38.1) & 37.3 (36.7, 38) & 37.5 (37, 38.1) & 37.4 (36.7, 38.1) \\
CRP (mg/L) & 7 & 48 (10, 118) & 43 (8, 110) & 85 (34, 160) & 91 (41, 180) \\
D dimer (µg/L) & 64 & 1100 (527, 2545) & 1028 (504, 2300) & 1950 (719, 8110) & 2100 (949, 4772) \\
LDH (U/L) & 18 & 244 (200, 322) & 237 (197, 302) & 300 (234, 422) & 338 (253, 492) \\
Leucocytes (x10\textasciicircum{9/L)} & 7 & 9.1 (6.7, 12.7) & 9.1 (6.7, 12.6) & 9.4 (6.5, 12.5) & 9.2 (6.4, 14) \\
Lymphocytes (x10\textasciicircum{9/L)} & 16 & 1.04 (0.66, 1.6) & 1.1 (0.7, 1.7) & 0.8 (0.54, 1.3) & 0.8 (0.51, 1.3) \\
Albumin (g/L) & 15 & 39 (36, 42) & 40 (36, 43) & 37 (34, 41) & 36 (33, 39) \\
Bicarbonate (mmol/L) & 45 & 23.6 (21, 26) & 24 (22, 26) & 23 (21, 26) & 22 (20, 25) \\
Creatinine (µmol/L) & 8 & 84 (66, 111) & 82 (65, 107) & 89 (66, 111) & 102 (75, 153) \\
Eosinophils (x10\textasciicircum{9/L)} & 26 & 0.03 (0, 0.1) & 0.03 (0.01, 0.1) & 0.03 (0, 0.1) & 0.01 (0, 0.1) \\
MCV (fL) & 7 & 90 (87, 94) & 90 (86, 94) & 90 (88, 94) & 91 (87, 96) \\
Monocytes (x10\textasciicircum{9/L)} & 30 & 0.67 (0.44, 0.95) & 0.68 (0.45, 0.95) & 0.59 (0.33, 0.83) & 0.61 (0.33, 0.94) \\
Neutrophils (x10\textasciicircum{9/L)} & 16 & 5.6 (2.2, 9) & 5.6 (2.2, 8.9) & 7 (4, 10.3) & 5.5 (1.8, 9.1) \\
Sodium (mmol/L) & 9 & 138 (135, 140) & 138 (135, 140) & 137 (133, 141) & 137 (134, 140) \\
Urea (mmol/L) & 9 & 6.5 (4.6, 9.7) & 6.2 (4.5, 9) & 7.4 (5.1, 11) & 9.6 (6.6, 15) \\
\end{tabular}}

\end{table}

\pagebreak

\begin{table}[h]
\ContinuedFloat
\caption{Continued.}
\footnotesize
\resizebox{\columnwidth}{!}{\begin{tabular}{lrrrrrrrrrrrrrr}
\toprule
 &  \% missing & All patients &   Discharged or dead &  In hospital & ICU admission \\
\midrule
\textbf{B. Validation data} &  n = 3252 & n = 2854 & n = 72 & n = 326 \\
Male sex & 0 & 56\% & 56\% & 46\% & 61\% \\
Age (years) & 0 & 71 (58, 80) & 69 (55, 79) & 72 (58, 82) & 79 (73, 85) \\
BMI (kg/m²) & 59 & 26 (23, 30) & 26 (23, 30) & 26 (24, 30) & 25 (22, 29) \\
HR (bpm) & 40 & 90 (78, 105) & 90 (78, 104) & 84 (75, 106) & 92 (78, 105) \\
SBP (mmHg) & 43 & 134 (119, 151) & 135 (120, 152) & 134 (122, 149) & 129 (110, 141) \\
RR (/min) & 43 & 20 (16, 24) & 20 (16, 24) & 20 (17, 26) & 23 (19, 27) \\
Saturation (\%) & 40 & 95.7 (94, 97.5) & 96 (94, 97.7) & 95.5 (94, 97) & 94.8 (92.3, 96.5) \\
Temperature (°C) & 42 & 37.3 (36.7, 38.1) & 37.3 (36.7, 38.1) & 37.3 (36.8, 38.1) & 37.2 (36.4, 38) \\
CRP (mg/L) & 9 & 57 (16, 124) & 54 (15, 120) & 76 (21, 169) & 80 (33, 159) \\
D dimer (µg/L) & 76 & 1060 (531, 2170) & 1013 (490, 2012) & 1080 (640, 2570) & 1495 (870, 3724) \\
LDH (U/L) & 22 & 247 (203, 334) & 242 (200, 317) & 281 (226, 390) & 315 (238, 489) \\
Leucocytes (x10\textasciicircum{9/L)} & 10 & 9.4 (6.6, 12.9) & 9.4 (6.6, 12.9) & 9.6 (7, 13.2) & 9.5 (6.8, 13.3) \\
Lymphocytes (x10\textasciicircum{9/L)} & 20 & 0.98 (0.62, 1.5) & 1 (0.64, 1.5) & 1.1 (0.72, 1.53) & 0.71 (0.48, 1.2) \\
Albumin (g/L) & 20 & 39 (35, 42) & 39 (35, 42) & 36 (32, 39) & 36 (32, 40) \\
Bicarbonate (mmol/L) & 50 & 23.5 (21, 26) & 23.6 (21, 26) & 22.8 (20, 27) & 22.8 (20, 25) \\
Creatinine (µmol/L) & 10 & 84 (66, 116) & 83 (65, 111) & 80 (58, 118) & 103 (74, 158) \\
Eosinophils (x10\textasciicircum{9/L)} & 27 & 0.03 (0.01, 0.1) & 0.03 (0.01, 0.1) & 0.03 (0.01, 0.09) & 0.03 (0, 0.04) \\
MCV (fL) & 10 & 90 (87, 94) & 90 (87, 94) & 92 (88, 95) & 91 (88, 96) \\
Monocytes (x10\textasciicircum{9/L)} & 30 & 0.67 (0.43, 0.98) & 0.67 (0.44, 0.98) & 0.57 (0.38, 1) & 0.58 (0.36, 0.97) \\
Neutrophils (x10\textasciicircum{9/L)} & 21 & 5.8 (2.4, 9.4) & 5.8 (2.4, 9.4) & 6.6 (3.3, 9.1) & 5.6 (2.2, 9.4) \\
Sodium (mmol/L) & 11 & 137 (134, 139) & 137 (134, 139) & 136 (133, 139) & 137 (133, 140) \\
Urea (mmol/L) & 11 & 6.9 (4.9, 10.4) & 6.6 (4.7, 9.8) & 7.3 (5.3, 11.8) & 10.3 (7.4, 17) \\
\bottomrule
\end{tabular}}
\end{table}

\end{landscape}

\begin{landscape}
\begin{table}[h]
\caption{Baseline characteristics of development and validation patient cohorts
of two hospitals with a full record of ICU admissions median (‘M’) and quartile
range (‘Q1’=first quartile; ‘Q3’=third quartile) are presented for all
continuous variables (continues to next page).}
\label{table:ch6-table2}
\footnotesize
\resizebox{\columnwidth}{!}{\begin{tabular}{lrrrrrrrrrrrrrr}
\toprule
                                       & &  \% missing & All patients &   Discharged or dead &  In hospital & ICU admission \\
\midrule
\textbf{A. Development data}           &            &    & n = 2633     & n = 2387        & n = 32     & n = 214 \\
Male sex                               &            & 0  & 58\%   & 57\%   & 63\%   & 68 \% \\
Age (years)                            &            & 0  & 71 (58, 80)  & 71 (57, 80)  & 80 (71, 85) & 68 (59, 74) \\
BMI (kg/m²)                            &            & 48 & 26 (23, 30)  & 26 (23, 30)  & 25 (22, 29) & 27 (24, 31) \\
HR (bpm)                               &            & 14 & 88 (77,100)  & 88 (77, 100) & 87 (77, 98) & 91 (79, 104)\\
SBP (mmHg)                             &            & 20 & 131 (116, 149)  & 131 (116 , 149)  & 140 (122 , 159)  & 129 (114 , 145)  \\
RR (/min)                              &            & 15 & 19 (15, 23)   & 19 (15, 23)   & 18 (16 , 22)   & 23 (18 , 27)   \\
Saturation (\%)                        &            & 10 & 95.(94, 97).3 & 95.(94, 97).4 & 96 (95, 97).5 & 95 (93.1, 96).7 \\
Temperature (°C)                       &            & 11 & 37.2 (36.7, 37.9) & 37.2 (36.7, 37.9) & 37.4 (36.7, 37.8) & 37.4 (36.7, 38.2) \\
CRP (mg/L)                             &            & 4  & (11, 120)  & 47 (10, 110)  & 82 (41, 175)  & 88 (20, 178) \\
D dimer (µg/L)                         &            & 79 & 1100 (463, 2700) & 1040 (460, 2453) & 3400 (2000, 5400) & 2000 (590, 4900) \\
LDH (U/L)                              &            & 11 & 239 (197, 317)  & 234 (194, 306)  & 270 (216, 324)  & 318 (234, 444)  \\
Leucocytes (x10\textasciicircum{}9/L)  &            & 4  & 9.5 (6.7, 13.2) & 9.4 (6.8, 13.1) & 12 (9.2, 16.1) & 10.3 (6.3, 14.7) \\
Lymphocytes (x10\textasciicircum{}9/L) &            & 11 & 1 (0.67, 1.6)  & 1 (0.7, 1.6) & 0.74 (0.55, 1.3)  & 1.1 (0.7, 1.5)  \\
Albumin (g/L)                          &            & 8  & 40 (36, 43)   & 40 (36, 43)   & 37 (34, 41)   & 38 (35, 42)   \\
Bicarbonate (mmol/L)                   &            & 66 & 23 (21, 25)   & 23 (21, 25)   & 24 (22, 26)   & 21 (19, 24)   \\
Creatinine (µmol/L)                    &            & 5  & 85 (67, 113)  & 84 (67, 112)  & 85 (64, 113)  & 93 (73, 125)  \\
Eosinophils (x10\textasciicircum{}9/L) &            & 27 & 0.1 (0.03, 0.1)  & 0.1 (0.03, 0.11) & 0.03 (0.03, 0.1) & 0.1 (0.03, 0.1)  \\
MCV (fL)                               &            & 5  & 91 (87, 95)   & 91 (87, 95)   & 92 (90, 97)   & 91 (88, 95)   \\
Monocytes (x10\textasciicircum{}9/L)   &            & 41 & 0.72 (0.47, 1) & 0.73 (0.48, 1) & 0.71 (0.56, 1.02) & 0.64 (0.39, 1)    \\
Neutrophils (x10\textasciicircum{}9/L) &            & 11 & 7.2 (4.8, 10.5) & 7.1 (4.8, 10.4) & 9.8 (6.6, 14.5) & 7.7 (5, 11.1) \\
Sodium (mmol/L)                        &            & 5  & 137 (134, 140)  & 137 (134, 140)  & 138 (134, 140)  & 137 (134, 139) \\
Urea (mmol/L)                          &            & 5  & 6.6 (4.7, 9.9)  & 6.6 (4.6, 9.8)  & 8.4 (6.3, 11.9) & 7.1 (5.1, 11)
\end{tabular}}
\end{table}

\pagebreak

\begin{table}[h]
\ContinuedFloat
\caption{Continued.}
\footnotesize
\resizebox{\columnwidth}{!}{\begin{tabular}{lrrrrrrrrrrrrrr}
\toprule
 &             &  \% missing & All patients &   Discharged or dead &  In hospital & ICU admission \\
\midrule
\textbf{B. Validation data}            &           & & n = 1466          & n = 1356           & n = 24      & n = 86 \\
Male sex                               &            & 0  & 58\%   & 58 \%   & 38\%   & 71\% \\
Age (years)                            &            & 0  & 71 (57, 80)  & 71 (56, 80)   & 80 (69, 85) & 70 (62, 76) \\
BMI (kg/m²)                            &            & 40 & 26 (23, 30) & 26 (23, 30)   & 25 (22, 31)   & 28 (24, 30) \\
HR (bpm)                               &            & 16 & 89 (77, 103)  & 89 (77, 103)  & 78 (73, 99)   & 89 (78, 103)  \\
SBP (mmHg)                             &            & 21 & 134 (119, 151)  & 134 (119, 151)  & 132 (111, 149)  & 141 (124, 157)  \\
RR (/min)                              &            & 17 & 19 (16, 23)   & 19 (16, 23)   & 19 (16, 22)   & 23 (19, 28)   \\
Saturation (\%)                        &            & 9  & 95.7 (94, 97.3) & 95.9 (94, 97.5) & 96.2 (94.5, 97)   & 93.8 (91.3, 95.7) \\
Temperature (°C)                       &            & 13 & 37.2 (36.7, 37.9) & 37.2 (36.7, 37.9) & 37.2 (36.7, 37.7) & 37.4 (37, 38)   \\
CRP (mg/L)                             &            & 5  & 49 (14, 115)  & 47 (13, 110)  & 77 (25, 135)  & 114 (40, 206)  \\
D dimer (µg/L)                         &            & 95 & 460 (210, 1275) & 440 (210, 1350) & 770 (770, 770)  & 913 (890, 937)  \\
LDH (U/L)                              &            & 13 & 238 (196, 314)  & 234 (195, 307)  & 254 (217, 288)  & 360 (234, 531)  \\
Leucocytes (x10\textasciicircum{}9/L)  &            & 6  & 9.7 (6.8, 13.2) & 9.7 (6.8, 13.2) & 12 (7.3, 15)   & 8.9 (6.7, 13.1) \\
Lymphocytes (x10\textasciicircum{}9/L) &            & 12 & 1 (0.65, 1.5)  & 1 (0.66 , 1.5)  & 1.1 (0.8, 1.33) & 0.8 (0.6, 1.55) \\
Albumin (g/L)                          &            & 9  & 40 (37, 42)   & 40 (37, 42)   & 37 (35, 38)   & 39 (36, 42)   \\
Bicarbonate (mmol/L)                   &            & 65 & 23 (21, 25)   & 23 (21, 25)   & 23 (21, 26)   & 22 (19, 25)   \\
Creatinine (µmol/L)                    &            & 6  & 86 (68, 119)  & 85 (68, 117)  & 80 (58, 122)  & 96 (77, 138)  \\
Eosinophils (x10\textasciicircum{}9/L) &            & 20 & 0.07 (0.03, 0.1)  & 0.07 (0.03, 0.11) & 0.05 (0.03, 0.1)  & 0.05 (0.03, 0.1)  \\
MCV (fL)                               &            & 6  & 91 (88, 95)   & 91 (88, 95)   & 94 (90, 97)   & 91 (87, 95)   \\
Monocytes (x10\textasciicircum{}9/L)   &            & 33 & 0.7 (0.47, 1)    & 0.7 (0.48, 1)    & 0.68 (0.48, 1.03) & 0.61 (0.4, 1)    \\
Neutrophils (x10\textasciicircum{}9/L) &            & 13 & 7.4 (4.8 , 11)   & 7.3 (4.7, 11)   & 9 (5.5, 13.1) & 7.4 (5  , 10.8) \\
Sodium (mmol/L)                        &            & 6  & 137 (134, 139)  & 137 (134, 139)  & 136 (133, 138)  & 135 (132, 138)  \\
Urea (mmol/L)                          &            & 7  & 6.7 (4.9 , 10)   & 6.6 (4.8, 9.9)  & 8 (4.7, 12.8) & 7.8 (5.9, 12.2) \\
\bottomrule
\end{tabular}}
\end{table}

\pagebreak



\end{landscape}

### Prediction of death {.unnumbered}

Patients who were admitted in the first month of the pandemic in the
Netherlands, that is, in March 2020, were at substantially increased risk of
death (Table \ref{table:ch6-table3}: multivariable OR 1.99; 95% CI 1.61 to
2.47). All models included this correction factor for the first month, to avoid
overestimation of risk after the first month of the pandemic. Consequently, to
avoid overestimation of the discriminative ability, we limited validation of
models in the development data to patients who were admitted from April 2020
onward.

D-dimer concentration in the blood, measured to detect thrombosis, was not
analysed in the regression analysis, because 64% and 76% were missing in the
development and validation data, respectively (Table \ref{table:ch6-table1}).
Based on a full model with restricted cubic splines of all potential variables,
we decided to transform all biomarkers and RR with the natural logarithm, while
keeping all other predictor effects linear. Some strong univariable associations
with death—for example of logarithmically transformed lymphocytes and creatinine
(Table \ref{table:ch6-table3}; Wald statistics 48 and 133, respectively)—were
very weak in multivariable analysis (Table \ref{table:ch6-table3}; Wald
statistics 0 and 4, respectively). The predictive ability of the resulting full
multivariable regression model was mainly driven by age, LDH, urea, RR, CRP,
Albumin, oxygen saturation and bicarbonate (ORs and Wald statistics in Table
\ref{table:ch6-table3}). A simple model—named COVID outcome prediction in the
emergency department (COPE)—with linear age and logarithmic transforms of RR,
CRP, LDH, albumin and urea captured most of the ability to predict death within
28 days (Table \ref{table:ch6-table3}; @fig-multivariable-effects). Based on
internal bootstrap validation, we applied a shrinkage factor of 0.93 to the
regression coefficients.

COPE showed good discrimination for predicting death in 4498 patients who were
admitted from April up to and including August 2020 in the first wave (online
supplemental Figure 1); AUC in 4 hospitals 0.85 (95% CI 0.81 to 0.88)); 0.81
(95% CI 0.71 to 0.91); 0.86 (95% CI 0.82 to 0.90); 0.85 (95% CI 0.81 to 0.88))
and, more importantly, in the validation sample of 3235 patients who were
admitted in the second wave from September up to and including December 2020
(@fig-temporal; AUC in four hospitals: 0.82 (95% CI 0.78 to 0.86); 0.82 (95% CI
0.74 to 0.90); 0.79 (95% CI 0.70 to 0.88); 0.83 (95% CI 0.79 to 0.86)). The
decrease in AUC over time was partly driven by less case mix
heterogeneity—expressed by a lower model-based AUC (mbc)—of second wave patients
(@fig-temporal; mbc in four hospitals: 0.81; 0.82; 0.81; 0.82) as compared with
first wave patients (online supplemental Figure 1); mbc in four hospitals 0.82;
0.85, 0.83, 0.84). COPE was well calibrated in second wave patients of each of
the four hospitals, both on average—expressed by hospital-specific calibration
intercepts: 0.08 (95% CI −0.15 to 0.30); −0.17 (95% CI −0.65 to 0.30); −0.01
(95% CI −0.40 to 0.39); −0.12 (95% CI −0.30 to 0.07)—and by predicted risk
levels—expressed by hospital-specific calibration slopes: 1.09 (95% CI 0.86 to
1.31); 0.90 (95% CI 0.49 to 1.32); 0.91 (95% CI 0.57 to 1.25); 0.97 (95% CI 0.79
to 1.14) (@fig-temporal).


```{r fig-multivariable-effects, cache=FALSE, echo=FALSE, fig.cap="Multivariable effects of continuous predictors of death within 28 days predictions of the logarithm of the odds by continuous predictor levels, with other predictor levels set to the median. Wald statistics are listed within each plot to express variable importance (higher is better). CRP, C reactive protein; LDH, lactic dehydrogenase; RR, respiratory rate.",out.width="80%"}
knitr::include_graphics("figures/ch6-Figure1.pdf")
```

\begin{landscape}
\begin{table}[h]
\caption{Univariable and multivariable associations between predictors and death
within 28 days OR with 95% CIs for separate variables (columns ‘univariable’),
for a model with all available predictors (columns ‘full model’) and for a model
with only the six strongest predictors (columns ‘selected model’)}
\label{table:ch6-table3}
\resizebox{\columnwidth}{!}{\begin{tabular}{llrrrrrrrrrrrrr}
\toprule
\multicolumn{1}{c}{} & \multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Univariable}} & \multicolumn{3}{c}{\textbf{Full model}} & \multicolumn{3}{c}{\textbf{Selected model}} & \multicolumn{1}{c}{\textbf{COPE}} \\
\cmidrule(rl){3-5} \cmidrule(rl){6-8} \cmidrule(rl){9-11} \cmidrule(rl){12-12}
Predictor                              & Contrast                    & OR          & 95\% CI &      Wald & OR         & 95\% CI & Wald & OR             & 95\% CI & Wald & OR   \\
\midrule
Month                                  & $\geq$ April vs $\leq$April & 2.57        & 2.16 to 3.05 & 114  & 1.99       & 1.61 to 2.47 & 39   & 2.06           & 1.68 to 2.52 & 49   & 1.96 \\
Sex                                    & male vs female             & 1.37         & 1.15 to 1.63 & 13   & 1.12       & 0.9 to 1.39 & 1    &                 &         &      &      &      \\
Age (years)                            & 80 vs 58                   & 3.07        & 2.63 to 3.58 & 201  & 3.16       & 2.56 to 3.91 & 113  & 2.95            & 2.42 to 3.6  & 112  & 2.74 \\
BMI (kg/m)                            & 35 vs 25                   & 1.07         & 0.89 to 1.28 & 1    & 1.09    & 0.9 to 1.34 & 1    &                  &         &      &      &      \\
HR (bpm)                               & 103 vs 78                  & 1.16        & 1.05 to 1.29 & 8    & 1.19       & 0.97 to 1.45 & 3    &                &         &      &      &      \\
SBP (mmHg)                             & 150 vs 118                 & 0.76        & 0.65 to 0.89 & 12   & 0.86       & 0.73 to 1.01 & 3    &                &         &      &      &      \\
RR (/min)                              & 23 vs 16                   & 1.98        & 1.77 to 2.21 & 150  & 1.63       & 1.34 to 1.99 & 24   & 1.91           & 1.63 to 2.23 & 64   & 1.82 \\
Saturation (\%)                        & 97.5 vs 94                 & 0.61        & 0.52 to 0.72 & 37   & 0.77       & 0.65 to 0.9  & 10   &                &         &      &      &      \\
Temperature (C)                       & 38 vs 37                   & 1.16         & 1.02 to 1.32 & 5    & 1.05       & 0.87 to 1.27 & 0    &                &         &      &      &      \\
CRP (mg/L)                             & 118 vs 10                  & 2.76        & 2.35 to 3.25 & 149  & 1.54       & 1.22 to 1.93 & 14   & 1.57           & 1.27 to 1.93 & 18   & 1.52 \\
LDH (U/L)                              & 322 vs 200                 & 2.17        & 1.99 to 2.36 & 309  & 1.83       & 1.62 to 2.06 & 99   & 1.85           & 1.66 to 2.05 & 127  & 1.77 \\
Leucocytes (x10\textasciicircum{}9/L)  & 12.7 vs 6.7                & 1.01        & 0.92 to 1.11 & 0    & 0.88       & 0.75 to 1.02 & 3    &                &         &      &      &      \\
Lymphocytes (x10\textasciicircum{}9/L) & 1.6 vs 0.66                & 0.67        & 0.6 to 0.75 & 48   & 1.03        & 0.9 to 1.19 & 0    &                 &         &      &      &      \\
Albumin (g/L)                          & 42 vs 36                   & 0.58        & 0.53 to 0.62 & 191  & 0.8        & 0.71 to 0.9  & 14   & 0.77           & 0.69 to 0.86 & 21   & 0.78 \\
Bicarbonate (mmol/L)                   & 25.9 vs 21.4               & 0.71        & 0.65 to 0.78 & 54   & 0.81       & 0.71 to 0.92 & 10   &                &         &      &      &      \\
Creatinine (Ámol/L)                    & 111 vs 66                  & 1.58        & 1.46 to 1.71 & 133  & 0.85       & 0.72 to 1    & 4    &                &         &      &      &      \\
Eosinophils (x10\textasciicircum{}9/L) & 0.1 vs 0.004               & 0.77        & 0.64 to 0.93 & 7    & 1.08       & 0.9 to 1.3  & 1    &                 &         &      &      &      \\
MCV (fL)                               & 94 vs 87                   & 1.18        & 1.08 to 1.29 & 14   & 1.1        & 0.99 to 1.23 & 3    &                &         &      &      &      \\
Monocytes (x10\textasciicircum{}9/L)   & 0.95 vs 0.44               & 0.84        & 0.74 to 0.96 & 7    & 1.08       & 0.87 to 1.35 & 1    &                &         &      &      &      \\
Neutrophils (x10\textasciicircum{}9/L) & 9 vs 2.2                   & 0.97        & 0.88 to 1.07 & 0    & 0.94       & 0.82 to 1.08 & 1    &                &         &      &      &      \\
Sodium (mmol/L)                        & 140 vs 135                 & 0.98        & 0.96 to 1.01 & 1    & 1.03       & 0.98 to 1.09 & 1    &                &         &      &      &      \\
Urea (mmol/L)                          & 9.7 vs 4.6                 & 2.48        & 2.24 to 2.76 & 291  & 1.79       & 1.43 to 2.24 & 26   & 1.61           & 1.42 to 1.83 & 53   & 1.56 \\
\bottomrule
\end{tabular}}
\end{table}
\end{landscape}


```{r fig-temporal, cache=FALSE, echo=FALSE, fig.show="hold", fig.cap="Temporal validation: Performance of COPE for predicting death in second wave patients calibration plots of patients who were admitted since September 2020 in four separate Dutch hospitals. n is number of patients; a=calibration intercept (0 is perfect); b=calibration slope (1 is perfect); c=AUC (0.5 is useless; 1 is perfect); mbc=model-based AUC. AUC, area under the curve; COPE, COVID-19 outcome prediction in the emergency department.", out.width="80%"}
knitr::include_graphics("figures/ch6-Figure2.pdf")
```

When stratifying second wave patients according to a mortality risk threshold
equal to the event rate (10%), COPE assigned high risk to 246/326 patients who
actually died (76% sensitivity, ie, 24% false negatives) and low risk to
2086/2926 patients who actually survived (71% specificity, ie, 29% false
positives). With a 5% risk threshold, the sensitivity increased to 93% while the
specificity decreased to 49%. Based on a 20% risk threshold, the sensitivity
decreased to 49% while the specificity increased to 89%.


### Prediction of need for ICU admission {.unnumbered}

The probability of being admitted to the ICU was decreasing with age after the
age of 70 (OR of age ≥80 vs 70–79: 0.17 (95% CI 0.10 to 0.30)), likely
reflecting the decision not to admit older patients to the ICU. When adjusting
for this decreasing age effect after the age of 70—by including a linear spline
with a knot at age 70 in the regression model (online supplemental Figure 2)—the
strongest predictors of death were also predictive of ICU admission within 28
days, but associations were generally weaker for the latter (table 4 vs table
3). In patients below the age of 70, admitted from April up to and including
August 2020, a model with the linear predictor of death calibrated to ICU
admission had similar discriminative ability to a model that refitted all the
predictor effects (AUC 0.71 for both models). For robustness, we implemented the
calibrated model, also adjusted for a linearly decreasing age effect after the
age of 70, and not the refitted model (calibration slope 0.60; 95% CI 0.49 to
0.70) into COPE for predicting ICU admission. To predict the need for ICU
admission of future patients over the age of 70 COPE ignores the decreasing age
effect after the age of 70, since the observed ICU admission rate is probably an
inaccurate estimate of the medical need for ICU admission. By fitting a linearly
decreasing age effect in patients over the age of 70 which is not applied when
predicting for future patients, predictions of ICU admission after the age of 70
are based on an extrapolation of the observed age effect on ICU admission in
patients below the age of 70. Due to the weaker predictor effects, the
discriminative ability of COPE was more moderate for predicting ICU admission
than for predicting death (online supplemental Figure 3); AUC in two hospitals:
0.66 (95% CI 0.58 to 0.74); 0.79 (95% CI 0.69 to 0.88)). Although COPE
significantly overestimated ICU admission in second wave patients
(@fig-temporal2; calibration intercept in two hospitals: −0.41 (95% CI −0.77 to
−0.05); −0.72 (95% CI −1.34 to −0.11)), it was better able to identify the
patients at high risk of needing ICU admission, as expressed by higher
discriminative ability (@fig-temporal2; AUC in two hospitals: 0.84 (95% CI 0.78
to 0.90); 0.81 (95% CI 0.66 to 0.95)) and substantially stronger predictor
effects (calibration slope in two hospitals: 1.55 (95% CI 1.03 to 2.06); 1.53
(95% CI 0.60 to 2.46)).

```{r fig-temporal2, cache=FALSE, echo=FALSE, fig.cap="Temporal validation: Performance of COPE for predicting ICU admission in second wave patients calibration plots of patients who were admitted since September 2020 in two separate Dutch hospitals. n is number of patients; a=calibration intercept (0 is perfect); b=calibration slope (1 is perfect); c=AUC (0.5 is useless; 1 is perfect); mbc=model-based AUC. AUC, area under the curve; COPE, COVID-19 outcome prediction in the emergency department; ICU, intensive care unit.",out.width="80%"}
knitr::include_graphics("figures/ch6-Figure3.pdf")
```
### Model presentation {.unnumbered}

The resulting COPE models for predicting death as well as need for ICU admission
within 28 days after hospital admission (formulas in Table
\ref{table:ch6-table5}) are implemented as a publicly accessible web-based
application
[https://mdmerasmusmc.shinyapps.io/COPE/](https://mdmerasmusmc.shinyapps.io/COPE/)
and as independent mobile apps (‘COPE Decision Support’). For optimal
transparency, the web and mobile applications include a detailed description of
the derivation of COPE (online supplemental file 1), descriptions of the data
that were used for development and validation of COPE, and calibration plots of
temporal validation in the separate hospitals. According to the Transparent
Reporting of a multivariable prediction model for Individual Prognosis Or
Diagnosis checklist (online supplemental table 1), all relevant items are
covered in this manuscript, except for the availability of data sets
[@collins2015transparent;@moons2015transparent]. The data that support the
findings of this study are available on request from the corresponding author.
The data are not publicly available due to data transfer agreements with each of
the contributing hospitals.

## Discussion {.unnumbered}

We developed COPE for prediction of in-hospital death and need for intensive
care when patients with suspected COVID-19 present at the ED. Developed using
patient data from the first wave of the pandemic, based on six quickly and
objectively obtainable predictors when entering the ED—age, RR, LDH, CRP,
albumin and urea—COPE discriminated well and was well calibrated in patients
admitted to hospitals in the second wave of the pandemic, both for predicting
in-hospital death and for ICU admission.

The clinical presentation of COVID-19 is broad and varies from asymptomatic to
critical disease. Some patients who initially have mild symptoms progress to
severe disease within 1 week [@cohen2020early]. In the ED physicians need to
identify high-risk patients—that is, those at high risk of deterioration and/or
death—requiring treatment in the ICU, intermediate-risk patients requiring
admission to the clinical ward, and low-risk patients who can potentially be
sent home. Since COPE is based on data that are routinely measured, or at least
readily available in the ED, it can act as a tool to support such decisions.
Hospitalised patients who are at high risk for mortality or need for ICU
admission should be more intensively watched, and when a high load of high-risk
patients occurs in the ED, this should be taken into account in the ICU capacity
planning. COPE does not explicitly define treatment decisions based on risk
thresholds, such as: send the patient home when the mortality risk is below a
risk threshold of x%, or: admit the patient to the ICU over a mortality risk
threshold of y%. These currently unavailable risk thresholds, and hence the
resulting treatment decisions, depend on a trade-off between benefits and harms
(including costs) of hospital or ICU admission [@gold1996cost]. Further research
is necessary to better understand the benefits and harms of hospital admission
and of ICU admission, for individual patients with COVID-19
[@ioannidis2011individualized]. Furthermore, treatment decisions may depend on
the availability of resources. The decision to admit a patient to the hospital,
or even to the ICU, may depend on the availability of hospital beds and ICU
beds. Likewise, the decision to send a low-risk patient home may depend on the
availability of relatives who are willing to care for the patient at home.
Although it is currently not possible to define explicit risk-based treatment
decisions for patients with COVID-19, the risk predictions provided by COPE can
be factored in by doctors, patients and relatives, when making decisions about
hospital or ICU admission.

We requested 19 large Dutch hospitals to supply anonymised retrospective data on
the cohorts of patients with COVID-19 who were admitted to their hospital. This
request for data was sent out very early in the pandemic and was greeted with
enthusiasm. Probably due to the enormous pressure on healthcare at that time,
four hospitals supplied useable data for the analysis. The contributing
hospitals were well spread over the Netherlands, with one in the west, two in
the south and one in the east of the country and are a mix of academic and large
teaching hospitals. we believe they are representative for healthcare in the
Netherlands. Although the consistently good performance of COPE across the
hospitals may support its generalisability to other countries, geographical
validation would be additionally reassuring, since the epidemic, and clinical
practice—for example, access to ICU or other enhanced care—for this novel
disease, may have substantial intercountry variability.

COPE was developed based on 5831 patients of whom 629 died within 28 days. This
effective sample size of 629 events was ample to start the development process
with a full model of 45 regression coefficients (14 events per variable), that
is, one binary predictor (sex) and 22 continuous predictors with 2 regression
coefficients— due to using non-linear terms—each [@steyerberg2015prediction]. To
prevent too extreme predictions of COPE in new data, we applied a shrinkage
factor to its regression coefficients, based on a bootstrap procedure with
backward selection starting from the full model [@harrell2015regression].


\begin{landscape}
\begin{table}[h]
\caption{Multivariable associations between predictors and ICU admission within 28 days OR with 95% CIs for separate
variables (columns ‘univariable’) and for a model with the six strongest predictors of death, corrected for a decreasing
probability of ICU admission after the age of 70 (columns ‘multivariable’)}
\label{table:ch6-table4}
\resizebox{\columnwidth}{!}{\begin{tabular}{llrrrrrrrrrrrrr}
\toprule
\multicolumn{1}{c}{} & \multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Univariable}} & \multicolumn{3}{c}{\textbf{Multivariable}} \\
\cmidrule(rl){3-5} \cmidrule(rl){6-8}
Predictor                    & Contrast                   & OR          & 95\% CI & Wald & OR            & 95\% CI &      Wald\\
\midrule
Month                        & $\geq$April vs $\leq$April & 2.06        & 1.51 to 2.81 & 21   & 1.63          & 1.16 to 2.28 & 8    \\
Age (years)                  & 80 vs 58                   & 1.96        & 1.47 to 2.62 & 21   & 1.76          & 1.32 to 2.35 & 15   \\
RR (/min)                    & 23 vs 16                   & 1.76        & 1.48 to 2.09 & 40   & 1.71          & 1.4 to 2.09 & 27   \\
CRP (mg/L)                   & 118 vs 10                  & 1.88        & 1.44 to 2.44 & 22   & 1.3           & 0.95 to 1.77 & 3    \\
LDH (U/L)                    & 322 vs 200                 & 1.73        & 1.52 to 1.98 & 66   & 1.44          & 1.25 to 1.67 & 24   \\
Albumin (g/L)                & 42 vs 36                   & 0.75        & 0.64 to 0.88 & 13   & 0.95          & 0.78 to 1.17 & 0    \\
Urea (mmol/L)                & 9.7 vs 4.6                 & 1.29        & 1.08 to 1.54 & 8    & 1.36          & 1.1 to 1.66 & 8    \\
Adjusted for:                &                            &             &         &      &               &         &      \\
Max{[}Age-70, 0{]} (years) * & 80 vs 58                   & 0.2         & 0.13 to 0.3  & 57   & 0.17          & 0.11 to 0.27 & 65   \\
\bottomrule
\end{tabular}}
\end{table}
\end{landscape}


\begin{table}[h]
\caption{COPE definition. COPE: COVID-19 outcome prediction in the emergency
department; CRP: C reactive protein; ICU: intensive care unit;
LDH: lactic dehydrogenase: RR: respiratory rate.}
\label{table:ch6-table5}
\begin{tabular}{lrr}
\toprule
Predictor & Minimum & Maximum \\
\midrule
Age (years) & 0 & 100 \\
RR (/min) & 10 & 60 \\
CRP (mg/L) & 1 & 500 \\
LDH (U/L) & 50 & 4000 \\
Albumin (g/L) & 10 & 60 \\
Urea (mmol/L) & 1 & 80 \\
\midrule
\multicolumn{3}{c}{
$\begin{aligned} \text{lp} = &-13.6 + 0.04575\times \text{age} + 1.654\times\log(\text{RR}) + \\ & 0.1688\times\log(\text{CRP}) + 1.197\times\log(\text{LDH}) − \\ & 1.585\times\log(\text{albumin}) + 0.5953\times\log(\text{urea})\end{aligned}$
} \\
\midrule
\multicolumn{3}{l}{\makecell{Probability of death \\ within 28 days $=\frac{1}{(1+exp(-\text{lp}))}$}} \\
\midrule
\multicolumn{3}{l}{\makecell{Probability of ICU admission \\ within 28 days $=\frac{1}{(1+exp(-(−0.08949+0.5970\times\text{lp})))}$}} \\
\bottomrule
\end{tabular}
\end{table}


Our explicit aim was to develop a score based on quickly and objectively
obtainable predictors at presentation at the ED. Consequently, pre-existing
comorbidities, the level of consciousness measured by the Glasgow Coma Scale,
and chest radiography results—although predictive for outcomes of patients with
COVID-19 in other studies—were not considered here
[@knight2020risk;@liang2020development;@king2020development]. Some predictors
were promising in univariable analysis, such as lymphocytes and creatinine, but
had negligible effects in multivariable analysis, because of strong correlations
with other, more important predictors. Other predictors, such as oxygen
saturation and bicarbonate, were significantly associated with death in
multivariable analysis, but were not selected into the final model, since our
explicit aim was to develop a simple model and the incremental value of these
predictors was minimal. To achieve this aim, we only selected the strongest
predictors—age, RR, LDH, CRP, albumin and urea—resulting in a parsimonious but
well-performing model.

We aimed to predict outcomes for all patients who present to the ED with
suspected COVID-19, regardless of actual hospital admission. Our data were
limited to patients who presented at the ED and were admitted to hospital,
because their outcomes were captured in the retrospective hospital database,
while outcomes of patients who were sent home were not captured in the
retrospective hospital database. Nevertheless, over 90% of the patients who
presented to the ED with suspected COVID-19 were admitted to hospital and it is
reasonable to assume that our predictions can be extrapolated to the less than
10% of patients who were sent home. Of note, the discriminative ability of our
model is probably better in all patients presenting to the ED, due to a more
heterogeneous case-mix: patients who were sent home are likely to have more
favourable predictor levels and more favourable outcomes than patients who are
admitted [@van_klaveren2016new].

Besides mortality, we aimed to predict the need for ICU admission. A limitation
of our study is that the need for ICU admission differs from the observed
decisions on ICU admission, and is inherently difficult to model, because
recorded ICU admissions express historical decisions at national, regional,
hospital or even intensivist level. As a robust solution, we exploited the
strong correlation between need for intensive care and death, by calibrating our
model for predicting death to the observed ICU admissions, adjusting for a
linear decrease with age after the age of 70. Hence, we assumed a linear
relationship between (the logarithm of the odds of) death and need for ICU
admission, and that all patients below the age of 70 needing intensive care were
actually admitted to the ICU, that is, the need for ICU admission is well
estimated by the observed decisions on ICU admission for patients below the age
of 70. The latter is reasonable given the sufficiency of ICU beds for Dutch
patients throughout the pandemic. The discriminative ability of this
recalibration approach was very similar to that of a model that refitted all
associations between COPE predictors and ICU admission. With temporal validation
in two separate hospitals, we showed that COPE discriminated very well between
patients at low and high risk of ICU admission and that the predicted
probability of ICU admission was well calibrated for the 20% highest-risk
patients (highest risk quintiles in [@fig-temporal2]). Nevertheless,
recalibration of COPE for predicting need for ICU admission to local
circumstances may be necessary.

The absence of external validation in our study— measuring the predictive
performance of COPE in hospitals that were not present in the development
data—may be considered a limitation of this study [@altman2000what]. However,
the combination of temporal validation—in second wave patients—and geographical
validation—in separate hospitals—is a strength of this study
[@austin2016geographic]. Although COPE already performed very well when
validated across time and space, future research should focus on analyses of
potential time trends not captured by the predictors—for example, changes in
mortality due to: improvements in treating patients with COVID-19; mutations of
COVID19; changes in patient case-mix or critical care capacity fluctuations
[@recovery_collaborative_group2021dexamethasone]
---potential changes in predictor effects in time (interactions between
predictors and time), and the impact of potential differences in patient case
mix and differences in clinical care in countries other than the Netherlands
(international validation). These case-mix and clinical care differences should
primarily affect calibration, requiring an update of the model intercept, but
not discrimination. The emergence of new COVID-19 variants with potentially
different mortality risk may especially require frequent analyses of the need
for model updating [@challen2021risk].

In conclusion, COPE, a simple tool based on six quickly and objectively
obtainable predictors in the ED, is well able to predict mortality and need for
ICU admission for patients who present to the ED with suspected COVID19. COPE
may support patients and doctors in decision making.

## Supplementary material {.unnumbered}

Supplementary material for this chapter is available online at
[https://doi.org/10.1016/j.ejca.2020.04.022](https://doi.org/10.1016/j.ejca.2020.04.022).

\clearpage\null\pagestyle{empty}
