\pagestyle{fancy}
# Teriparatide vs bisphosphonates: A treatment effect heterogeneity study {#sec-osteoporosis-appendix}

\pagebreak

## Cohort definitions {#sec-osteoporosis-appendix-cohorts}

### Teriparatide {.unnumbered}

\textbf{Cohort Entry Events} \par
People with continuous observation of 365 days
before event may enter the cohort when observing any of the following:

* drug exposures of Teriparatide (Table \ref{table:ch7-appendix-teriparatide})

Limit cohort entry events to the earliest event per person.

**Inclusion Criteria:**

* [Inclusion] Established osteoporosis: Any fracture of the hip, wrist, spine, shoulder/humerus

  * Entry events having at least 1 condition occurrence of osteoporosis (Table \ref{table:ch7-appendix-osteoporosis}), starting in the 365 days prior to cohort entry start date

* [Inclusion] Women above the age of 50
  * Entry events with the following event criteria: who are female > 50 years old
* [Exclusion] No anti-osteoporosis drugs 1 year prior: SERMs (raloxifene, bazedoxifene), denosumab, abaloparatide (US), romosozumab (US)
  * Entry events having no drug exposures of anti-osteoporosis drugs, other than teriparatide or bisphosphonates (Table \ref{table:ch7-appendix-antiosteoporosis-drugs}), starting in the 365 days prior to cohort entry start date
* [Exclusion] No oral bisphosphonate treatment
  * Entry events having no drug exposures of oral bisphosphonates (Table \ref{table:ch7-appendix-bisphosphonates}), starting between 365 days before and 0 days before cohort entry start date.
* [Exclusion] No history of breast cancer
  * Entry events having no condition occurrences of breast cancer (Table \ref{table:ch7-appendix-breast-cancer}), starting any time prior to cohort entry start date.
* [Exclusion] No Paget's disease
  * Entry events having no condition occurrences of Paget's disease of bone (Table \ref{table:ch7-appendix-paget}), starting any time prior to cohort entry start date
  
**Cohort Exit** \par
The person exits the cohort at the end of continuous observation.

**Cohort Eras** \par
Entry events will be combined into cohort eras if they are within 0 days of each other.


### Oral bisphosphonates {.unnumbered}

\textbf{Cohort Entry Events} \par
People with continuous observation of 365 days
before event may enter the cohort when observing any of the following:

* drug exposures of oral bisphosphonates (Table \ref{table:ch7-appendix-bisphosphonates})

Limit cohort entry events to the earliest event per person.

**Inclusion Criteria:**

* [Inclusion] Established osteoporosis: Any fracture of the hip, wrist, spine, shoulder/humerus
  ** Entry events having at least 1 condition occurrence of '[PR] Osteoporosis', starting in the 365 days prior to cohort entry start date.
* [Inclusion] Women above the age of 50
  * Entry events with the following event criteria: who are female > 50 years old.
* [Exclusion] No anti-osteoporosis drugs 1 year prior: SERMs(raloxifene, bazedoxifene), denosumab, abaloparatide (US), romosozumab (US)
  * Entry events having no drug exposures of anti-osteoporosis drugs, other than teriparatide or bisphosphonates (\ref{table:ch7-appendix-antiosteoporosis-drugs}), starting in the 365 days prior to cohort entry start date.
* [Exclusion] No teriparatide treatment
  * Entry events having no drug exposures of teriparatide (Table \ref{table:ch7-appendix-teriparatide}), starting in the 365 days prior to cohort entry start date.
* [Exclusion] No history of breast cancer
  * Entry events having no condition occurrences of breast cancer (Table \ref{table:ch7-appendix-breast-cancer}), starting any time prior to cohort entry start date.
* [Exclusion] No Paget's disease
  * Entry events having no condition occurrences of Paget's disease of bone (Table \ref{table:ch7-appendix-paget}), starting any time prior to cohort entry start date

**Cohort Exit** \par
The person exits the cohort at the end of continuous observation.

**Cohort Eras** \par
Entry events will be combined into cohort eras if they are within 0 days of each other.


### Hip fracture {.unnumbered}

\textbf{Cohort Entry Events} \par
People may enter the cohort when observing any of the following:

* condition occurence of hip fracture (Table \ref{table:ch7-appendix-hip-fracture})

Restrict entry events to having at most 0 condition occurrences of hip fracture,
starting in the 180 days prior to cohort entry start date. Limit qualifying
entry events to the all events per person.

**Cohort Exit** \par
The person exits the cohort at the end of continuous observation.

**Cohort Eras** \par
Entry events will be combined into cohort eras if they are within 0 days of each other.

### Major osteoporotic fracture {.unnumbered}

\textbf{Cohort Entry Events} \par
People with continuous observation of 180 days before event may enter the cohort
when observing any of the following:

* condition occurrences of hip fracture (Table \ref{table:ch7-appendix-hip-fracture}).
* condition occurrences of vertebral fracture (Table \ref{table:ch7-appendix-vertebral-fracture}).
* condition occurrences of wrist-forearm-proximal humerus fractures (Table \ref{table:ch7-appendix-wrist}).

Restrict entry events to with all of the following criteria:

* having at most 0 condition occurrences of hip fracture (Table \ref{table:ch7-appendix-hip-fracture}), starting in the 180 days prior to cohort entry start date.
* having at most 0 condition occurrences of vertebral fracture (Table \ref{table:ch7-appendix-vertebral-fracture}), starting in the 180 days prior to cohort entry start date.
* having at most 0 condition occurrences of wrist-forearm-proximal humerus fractures (Table \ref{table:ch7-appendix-wrist}), starting in the 180 days prior to cohort entry start date.

Limit these restricted entry events to the earliest event per person.

**Cohort Exit** \par
The person exits the cohort at the end of continuous observation.

**Cohort Eras** \par
Entry events will be combined into cohort eras if they are within 0 days of each other.

### Vertebral fracture {.unnumbered}

\textbf{Cohort Entry Events} \par
People may enter the cohort when observing any of the following:

* condition occurrences of vertebral fracture (Table \ref{table:ch7-appendix-vertebral-fracture}).

Restrict entry events to having at most 0 condition occurrences of vertebral
fracture (Table \ref{table:ch7-appendix-vertebral-fracture}), starting in the 180 days prior to cohort entry
start date. Limit these restricted entry events to the earliest event per
person.

**Cohort Exit** \par
The person exits the cohort at the end of continuous observation.

**Cohort Eras** \par
Entry events will be combined into cohort eras if they are within 0 days of each other.


\begin{table}[H]
\caption{Concepts used to define anti-osteoporosis drugs (other than teriparatide or bisphosphonates).}
\label{table:ch7-appendix-antiosteoporosis-drugs}
\resizebox{\textwidth}{!}{\begin{tabular}{llllll}
\toprule
Concept Id & Concept name & Domain & Vocabulary & Excluded & Decendants \\
\midrule
1594148 & abaloparatide & Drug & RxNorm & NO & YES \\
44506794 & bazedoxifene & Drug & RxNorm & NO & YES \\
40222444 & denosumab & Drug & RxNorm & NO & YES \\
1513103 & Raloxifene & Drug & RxNorm & NO & YES \\
1511251 & romosozumab & Drug & RxNorm & NO & YES \\
19000815 & Strontium & Drug & RxNorm & NO & YES\\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[H]
\caption{Concepts used to define breast cancer.}
\label{table:ch7-appendix-breast-cancer}
\resizebox{\textwidth}{!}{\begin{tabular}{llllll}
\toprule
Concept Id & Concept name & Domain & Vocabulary & Excluded & Decendants \\
\midrule
4112853 & Malignant tumor of breast & Condition & SNOMED & NO & YES \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[H]
\caption{Concepts used to  define oral bisphosphonates.}
\label{table:ch7-appendix-bisphosphonates}
\resizebox{\textwidth}{!}{\begin{tabular}{llllll}
\toprule
Concept Id & Concept name & Domain & Vocabulary & Excluded & Decendants \\
\bottomrule
1557272 & Alendronate & Drug & RxNorm & NO & YES \\
21604152 & alendronic acid & Drug & ATC & NO & YES \\
1512480 & Ibandronate & Drug & RxNorm & NO & YES \\
21604154 & ibandronic acid & Drug & ATC & NO & YES \\
1516800 & Risedronate & Drug & RxNorm & NO & YES \\
21604155 & risedronic acid & Drug & ATC & NO & YES \\
1524674 & zoledronic acid & Drug & RxNorm & NO & YES \\
21604156 & zoledronic acid & Drug & ATC & NO & YES \\
40241457 & zoledronic acid 0.04 MG/ML {[}Zometa{]} & Drug & RxNorm & YES & YES \\
1524677 & zoledronic acid 0.8 MG/ML & Drug & RxNorm & YES & YES \\
21114423 & zoledronic acid 4 MG {[}Zometa{]} & Drug & RxNorm Extension & YES & YES \\
42956218 & zoledronic acid Injectable Solution {[}Zometa{]} & Drug & RxNorm Extension & YES & YES \\
42956220 & zoledronic acid Injectable Solution {[}ZOMETA READY{]} & Drug & RxNorm Extension & YES & YES \\
41112021 & zoledronic acid Injectable Suspension {[}Zometa{]} & Drug & RxNorm Extension & YES & YES \\
1593366 & zoledronic acid Injection {[}Zometa{]} & Drug & RxNorm & YES & YES \\
36787242 & zoledronic acid Intravenous Solution {[}Zometa{]} & Drug & RxNorm Extension & YES & YES \\
36882768 & zoledronic acid Prefilled Syringe {[}Zometa{]} & Drug & RxNorm Extension & YES & YES \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[H]
\caption{Concepts used to define osteoporosis}
\label{table:ch7-appendix-osteoporosis}
\resizebox{\textwidth}{!}{\begin{tabular}{llllll}
\toprule
Concept Id & Concept name & Domain & Vocabulary & Excluded & Decendants \\
\midrule
4138275 & Fracture dislocation of wrist joint & Condition & SNOMED & NO & YES \\
45763653 & Fracture of bone of hip region & Condition & SNOMED & NO & YES \\
4218884 & Fracture of carpal bone & Condition & SNOMED & NO & YES \\
442619 & Fracture of humerus & Condition & SNOMED & NO & YES \\
4174520 & Fracture of vertebral column & Condition & SNOMED & NO & YES \\
80502 & Osteoporosis & Condition & SNOMED & NO & YES \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[H]
\caption{Concepts used to define Paget's disease.}
\label{table:ch7-appendix-paget}
\resizebox{\textwidth}{!}{\begin{tabular}{llllll}
\toprule
Concept Id & Concept name & Domain & Vocabulary & Excluded & Decendants \\
\midrule
4003461 & Paget's disease-clavicle & Condition & SNOMED & NO & YES \\
4002990 & Paget's disease-femur & Condition & SNOMED & NO & YES \\
4103167 & Pagets disease - hip & Condition & SNOMED & NO & YES \\
4001314 & Paget's disease-lumbar spine & Condition & SNOMED & NO & YES \\
4001445 & Paget's disease of humerus & Condition & SNOMED & NO & YES \\
4001981 & Paget's disease of pelvis & Condition & SNOMED & NO & YES \\
4001632 & Paget's disease of skull & Condition & SNOMED & NO & YES \\
4001313 & Paget's disease-thoracic spine & Condition & SNOMED & NO & YES \\
4003464 & Paget's disease-tibia & Condition & SNOMED & NO & YES \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[H]
\caption{Concepts used to define treatment with teriparatide.}
\label{table:ch7-appendix-teriparatide}
\resizebox{\textwidth}{!}{\begin{tabular}{lllllll}
\toprule
Concept Id & Concept name & Domain & Vocabulary & Excluded & Decendants & Mapped \\
\midrule
1521987 & Teriparatide & Drug & RxNorm & NO & YES & NO \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[H]
\caption{Concepts used to define hip fracture.}
\label{table:ch7-appendix-hip-fracture}
\resizebox{\textwidth}{!}{\begin{tabular}{llllll}
\toprule
Concept Id & Concept Name & Domain & Vocabulary & Excluded & Descendants \\
\midrule
433856 & Fracture of neck of femur & Condition & SNOMED & NO & YES \\
4167913 & Open fracture of femur, subcapital & Condition & SNOMED & YES & YES \\
436248 & Open fracture of neck of femur & Condition & SNOMED & YES & YES \\
4012436 & Open fracture of proximal femur, pertrochanteric & Condition & SNOMED & YES & YES \\
4015976 & Open fracture proximal femur,subcapital, Garden grade unspec & Condition & SNOMED & YES & YES \\
4015975 & Open fracture proximal femur, transepiphyseal & Condition & SNOMED & YES & YES \\
4230399 & Closed fracture of hip & Condition & SNOMED & NO & YES \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[H]
\caption{Concepts used to define vertebral fracture.}
\label{table:ch7-appendix-vertebral-fracture}
\resizebox{\textwidth}{!}{\begin{tabular}{llllll}
\toprule
Concept Id & Concept Name & Domain & Vocabulary & Excluded & Descendants \\
\midrule
36211121 & Lumbar vertebral fracture & Condition & MedDRA & NO & YES \\
36211124 & Thoracic vertebral fracture & Condition & MedDRA & NO & YES \\
45757320 & Osteoporotic fracture of vertebra & Condition & SNOMED & NO & YES \\
4013604 & Closed fracture lumbar vertebra & Condition & SNOMED & NO & YES \\
4013596 & Closed fracture thoracic vertebra & Condition & SNOMED & NO & YES \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[H]
\caption{Concepts used to define wrist/forearm/proximal humerus fractures.}
\label{table:ch7-appendix-wrist}
\resizebox{\textwidth}{!}{\begin{tabular}{llllll}
\toprule
Concept Id & Concept Name & Domain & Vocabulary & Excluded & Descendants \\
\midrule
440230 & Closed fracture of proximal humerus, anatomical neck & Condition & SNOMED & NO & YES \\
441699 & Closed fracture proximal humerus, greater tuberosity & Condition & SNOMED & NO & YES \\
4013920 & Closed fracture proximal humerus, neck & Condition & SNOMED & NO & YES \\
36211130 & Forearm fracture & Condition & MedDRA & NO & YES \\
432472 & Open fracture of proximal humerus, anatomical neck & Condition & SNOMED & NO & YES \\
437690 & Open fracture proximal humerus, greater tuberosity & Condition & SNOMED & NO & YES \\
4009431 & Open fracture proximal humerus, neck & Condition & SNOMED & NO & YES \\
36211137 & Wrist fracture & Condition & MedDRA & NO & YES \\
\bottomrule
\end{tabular}}
\end{table}



## Patient characteristics

\begin{landscape}
\begin{table}[H]
\caption{Characteristics of patients in CCAE receiving teriparatide or bisphosphonates before and after matching on the propensity score.}
\label{table:ch7-appendix-ccae}
\resizebox{\columnwidth}{!}{\begin{tabular}{lrrrrrr}
\toprule
& \multicolumn{3}{c}{Before matching} & \multicolumn{3}{c}{After matching} \\
\cmidrule(rl){2-4} \cmidrule(rl){5-7} 
Characteristic & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff \\
\midrule
\textbf{Medical history: General} &  &  &  &  &  &  \\
Attention deficit   hyperactivity disorder & 1.2 & 0.5 & 0.08 & 1.2 & 1.2 & 0 \\
Chronic obstructive lung   disease & 11.3 & 6.3 & 0.18 & 10.8 & 10.9 & -0.01 \\
Crohn's disease & 1.8 & 0.7 & 0.09 & 1.8 & 1.8 & 0 \\
Depressive disorder & 16.6 & 9 & 0.23 & 16 & 17.1 & -0.03 \\
Diabetes mellitus & 0.2 & 0.1 & 0.02 & 0.2 & 0.3 & -0.01 \\
Gastroesophageal reflux   disease & 27.6 & 14.8 & 0.32 & 26.4 & 27.5 & -0.02 \\
Gastrointestinal   hemorrhage & 3.3 & 1.8 & 0.1 & 3.3 & 3.4 & -0.01 \\
Human immunodeficiency   virus infection & 0.2 & 0.1 & 0.02 & 0.2 & 0.2 & -0.01 \\
Hyperlipidemia & 41.1 & 35.8 & 0.11 & 39.7 & 40.6 & -0.02 \\
Hypertensive disorder & 0.5 & 0.3 & 0.03 & 0.5 & 0.5 & 0 \\
Obesity & 7.1 & 4.3 & 0.12 & 6.8 & 7.5 & -0.03 \\
Osteoarthritis & 14.9 & 8.3 & 0.21 & 14.3 & 14.5 & -0.01 \\
Pneumonia & 1.8 & 0.7 & 0.1 & 1.6 & 1.9 & -0.02 \\
Psoriasis & 2.7 & 1.6 & 0.08 & 2.6 & 2.6 & 0 \\
Rheumatoid arthritis & 10.1 & 5.1 & 0.19 & 9.8 & 10.2 & -0.01 \\
Schizophrenia & 0.1 & 0.1 & 0.01 & 0.1 & 0.1 & 0 \\
Ulcerative colitis & 1.8 & 0.8 & 0.08 & 1.7 & 1.6 & 0.01 \\
Urinary tract infectious   disease & 33.5 & 21.8 & 0.26 & 31.9 & 32.7 & -0.02 \\
Viral hepatitis C & 0.8 & 0.3 & 0.06 & 0.8 & 0.7 & 0.01
\end{tabular}}
\end{table}

\begin{table}[H]
\ContinuedFloat
\caption{Continued.}
\resizebox{\columnwidth}{!}{\begin{tabular}{lrrrrrr}
\toprule
& \multicolumn{3}{c}{Before matching} & \multicolumn{3}{c}{After matching} \\
\cmidrule(rl){2-4} \cmidrule(rl){5-7} 
Characteristic & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff \\
\midrule
\textbf{Medical history: Cardiovascular disease} &  &  &  &  &  &  \\
Atrial fibrillation & 3.1 & 1.6 & 0.1 & 3 & 3.1 & 0 \\
Cerebrovascular disease & 1.9 & 1.1 & 0.07 & 1.8 & 1.8 & 0 \\
Coronary   arteriosclerosis & 5.6 & 3.4 & 0.11 & 5.4 & 5.4 & 0 \\
Heart disease & 1.8 & 0.8 & 0.09 & 1.8 & 1.6 & 0.01 \\
Heart failure & 1.1 & 0.4 & 0.09 & 1.1 & 1.1 & 0 \\
Ischemic heart disease & 0.2 & 0.1 & 0.02 & 0.2 & 0.2 & 0 \\
Pulmonary embolism & 0.4 & 0.1 & 0.05 & 0.4 & 0.4 & 0 \\
Venous thrombosis & 1.2 & 0.5 & 0.07 & 1 & 1.2 & -0.01 \\
\textbf{Medical history: Neoplasms} &  &  &  &  &  &  \\
Malignant lymphoma & 0.5 & 0.3 & 0.02 & 0.5 & 0.5 & 0 \\
Malignant neoplastic   disease & 0.4 & 0.3 & 0.02 & 0.4 & 0.4 & 0 \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[p]
\caption{Characteristics of patients in MDCR receiving teriparatide or bisphosphonates before and after matching on the propensity score.}
\label{table:ch7-appendix-mdcr}
\resizebox{\columnwidth}{!}{\begin{tabular}{lrrrrrr}
\toprule
& \multicolumn{3}{c}{Before matching} & \multicolumn{3}{c}{After matching} \\
\cmidrule(rl){2-4} \cmidrule(rl){5-7} 
Characteristic & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff \\
\midrule
\textbf{Medical history: General} &  &  &  &  &  &  \\
Attention deficit hyperactivity disorder & 0.3 & 0.1 & 0.04 & 0.3 & 0.2 & 0.02 \\
Chronic obstructive lung disease & 25.6 & 14.4 & 0.28 & 23.9 & 25.3 & -0.03 \\
Crohn's disease & 1.3 & 0.4 & 0.09 & 1.3 & 1.2 & 0.01 \\
Depressive disorder & 14.8 & 6.5 & 0.27 & 14.3 & 15.5 & -0.03 \\
Diabetes mellitus & 0.1 & 0 & 0.01 & 0.1 & 0.1 & 0 \\
Gastroesophageal reflux disease & 33.8 & 14.8 & 0.45 & 31.8 & 34.5 & -0.06 \\
Gastrointestinal hemorrhage & 7.3 & 3.3 & 0.18 & 6.7 & 7.1 & -0.01 \\
Human immunodeficiency virus infection & 0 & 0 & -0.01 & 0 & 0 & -0.02 \\
Hyperlipidemia & 47.5 & 33.5 & 0.29 & 46.4 & 48.8 & -0.05 \\
Obesity & 5.6 & 3.1 & 0.12 & 5.6 & 6.1 & -0.02 \\
Osteoarthritis & 27.1 & 14.7 & 0.31 & 25.3 & 26.9 & -0.04 \\
Pneumonia & 2.5 & 0.8 & 0.13 & 2.4 & 2.4 & 0 \\
Psoriasis & 2.6 & 1.4 & 0.09 & 2.6 & 3 & -0.03 \\
Rheumatoid arthritis & 11.6 & 5.2 & 0.23 & 11.1 & 11.4 & -0.01 \\
Schizophrenia & 0.1 & 0.1 & 0.01 & 0.1 & 0.1 & 0 \\
Ulcerative colitis & 1.9 & 0.7 & 0.11 & 1.8 & 1.7 & 0 \\
Urinary tract infectious disease & 38.1 & 20.5 & 0.39 & 36.2 & 38.2 & -0.04 \\
Viral hepatitis C & 0.3 & 0.1 & 0.04 & 0.4 & 0.3 & 0 \\
\end{tabular}}
\end{table}

\begin{table}[p]
\ContinuedFloat
\caption{Continued.}
\resizebox{\columnwidth}{!}{\begin{tabular}{lrrrrrr}
\toprule
& \multicolumn{3}{c}{Before matching} & \multicolumn{3}{c}{After matching} \\
\cmidrule(rl){2-4} \cmidrule(rl){5-7} 
Characteristic & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff \\
\midrule
\textbf{Medical history: Cardiovascular disease} &  &  &  &  &  &  \\
Atrial fibrillation & 15.4 & 9.3 & 0.19 & 14.6 & 15.1 & -0.01 \\
Cerebrovascular disease & 8.4 & 4 & 0.18 & 7.8 & 8.1 & -0.01 \\
Coronary arteriosclerosis & 19.1 & 11 & 0.23 & 17.8 & 18.3 & -0.01 \\
Heart disease & 4.4 & 1.8 & 0.15 & 4 & 4.3 & -0.02 \\
Heart failure & 2.8 & 1.2 & 0.12 & 2.7 & 3 & -0.02 \\
Ischemic heart disease & 0.7 & 0.4 & 0.04 & 0.6 & 0.7 & -0.01 \\
Pulmonary embolism & 0.6 & 0.2 & 0.06 & 0.5 & 0.6 & -0.01 \\
Venous thrombosis & 1.7 & 0.9 & 0.08 & 1.5 & 1.9 & -0.03 \\
\textbf{Medical history: Neoplasms} &  &  &  &  &  &  \\
Malignant lymphoma & 0.7 & 0.6 & 0.02 & 0.8 & 1 & -0.02 \\
Malignant neoplastic disease & 0.7 & 0.4 & 0.04 & 0.6 & 0.6 & -0.01 \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[p]
\caption{Characteristics of patients in Optum-DOD receiving teriparatide or bisphosphonates before and after matching on the propensity score.}
\label{table:ch7-appendix-optum-dod}
\resizebox{\columnwidth}{!}{\begin{tabular}{lrrrrrr}
\toprule
& \multicolumn{3}{c}{Before matching} & \multicolumn{3}{c}{After matching} \\
\cmidrule(rl){2-4} \cmidrule(rl){5-7} 
Characteristic & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff \\
\midrule
\textbf{Medical history: General} &  &  &  &  &  &  \\
Attention deficit   hyperactivity disorder & 0.8 & 0.3 & 0.07 & 0.8 & 0.9 & -0.01 \\
Chronic obstructive lung   disease & 22.6 & 15.3 & 0.19 & 22 & 23.2 & -0.03 \\
Crohn's disease & 1.4 & 0.6 & 0.08 & 1.3 & 1.4 & -0.01 \\
Depressive disorder & 21.8 & 11.5 & 0.28 & 20.4 & 22.1 & -0.04 \\
Diabetes mellitus & 0.3 & 0.3 & 0.01 & 0.3 & 0.4 & -0.02 \\
Gastroesophageal reflux   disease & 33.2 & 17.6 & 0.36 & 31 & 32.9 & -0.04 \\
Gastrointestinal   hemorrhage & 6 & 3 & 0.14 & 5.6 & 5.7 & -0.01 \\
Human immunodeficiency   virus infection & 0.2 & 0.1 & 0.02 & 0.2 & 0.2 & 0.01 \\
Hyperlipidemia & 56.2 & 53 & 0.06 & 54.4 & 55.7 & -0.03 \\
Obesity & 11.3 & 9.1 & 0.07 & 11.5 & 12.3 & -0.02 \\
Osteoarthritis & 30.6 & 20.5 & 0.23 & 29 & 30.1 & -0.02 \\
Pneumonia & 4.5 & 2.9 & 0.08 & 4.5 & 4.9 & -0.02 \\
Psoriasis & 2.8 & 1.7 & 0.07 & 2.6 & 2.8 & -0.02 \\
Rheumatoid arthritis & 13.7 & 6.5 & 0.24 & 13 & 13.5 & -0.01 \\
Schizophrenia & 0.3 & 0.3 & 0 & 0.3 & 0.3 & 0 \\
Ulcerative colitis & 1.8 & 0.8 & 0.09 & 1.7 & 1.6 & 0.01 \\
Urinary tract infectious   disease & 41.1 & 27.6 & 0.29 & 39 & 40.9 & -0.04 \\
Viral hepatitis C & 1.1 & 0.5 & 0.06 & 1.1 & 1.1 & 0 \\
\end{tabular}}
\end{table}

\begin{table}[p]
\ContinuedFloat
\caption{Continued.}
\resizebox{\columnwidth}{!}{\begin{tabular}{lrrrrrr}
\toprule
& \multicolumn{3}{c}{Before matching} & \multicolumn{3}{c}{After matching} \\
\cmidrule(rl){2-4} \cmidrule(rl){5-7} 
Characteristic & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff \\
\midrule
\textbf{Medical history: Cardiovascular disease} &  &  &  &  &  &  \\
Atrial fibrillation & 10.2 & 7 & 0.11 & 9.5 & 10.1 & -0.02 \\
Cerebrovascular disease & 5.4 & 3.4 & 0.1 & 5 & 5.3 & -0.01 \\
Coronary   arteriosclerosis & 11 & 6.6 & 0.16 & 10.2 & 10.7 & -0.02 \\
Heart disease & 4.6 & 2.6 & 0.11 & 4.3 & 4.4 & -0.01 \\
Heart failure & 4.8 & 3.1 & 0.09 & 4.9 & 5.3 & -0.02 \\
Ischemic heart disease & 0.4 & 0.3 & 0.03 & 0.4 & 0.4 & -0.01 \\
Pulmonary embolism & 1 & 0.6 & 0.04 & 1 & 1.1 & -0.01 \\
Venous thrombosis & 1.4 & 0.7 & 0.07 & 1.2 & 1.4 & -0.01 \\
\textbf{Medical history: Neoplasms} &  &  &  &  &  &  \\
Malignant lymphoma & 0.3 & 0.3 & 0 & 0.3 & 0.5 & -0.03 \\
Malignant neoplastic   disease & 0.4 & 0.3 & 0.02 & 0.3 & 0.5 & -0.02 \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[p]
\caption{Characteristics of patients in Optum-EHR receiving teriparatide or bisphosphonates before and after matching on the propensity score.}
\label{table:ch7-appendix-optum-ehr}
\resizebox{\columnwidth}{!}{\begin{tabular}{lrrrrrr}
\toprule
& \multicolumn{3}{c}{Before matching} & \multicolumn{3}{c}{After matching} \\
\cmidrule(rl){2-4} \cmidrule(rl){5-7} 
Characteristic & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff \\
\midrule
\textbf{Medical history: General} &  &  &  &  &  &  \\
Acute respiratory   disease & 0.1 & 0.1 & 0 & 0.1 & 0.1 & 0.01 \\
Attention deficit   hyperactivity disorder & 0.6 & 0.4 & 0.03 & 0.6 & 0.6 & 0 \\
Chronic obstructive lung   disease & 16.9 & 14 & 0.08 & 16.6 & 17.1 & -0.01 \\
Crohn's disease & 1.3 & 0.7 & 0.06 & 1.2 & 1.1 & 0.01 \\
Depressive disorder & 22 & 17 & 0.13 & 21.2 & 21.7 & -0.01 \\
Diabetes mellitus & 1.5 & 1.3 & 0.02 & 1.4 & 1.4 & 0.01 \\
Gastroesophageal reflux   disease & 31.4 & 24.9 & 0.15 & 30 & 29.9 & 0 \\
Gastrointestinal   hemorrhage & 3.7 & 2.6 & 0.06 & 3.5 & 3.6 & 0 \\
Human immunodeficiency   virus infection & 0.1 & 0.1 & -0.01 & 0.1 & 0.2 & -0.02 \\
Hyperlipidemia & 44.6 & 50.8 & -0.12 & 43.4 & 42 & 0.03 \\
Hypertensive disorder & 3.5 & 3.3 & 0.01 & 3.4 & 3.4 & 0 \\
Obesity & 10.3 & 9.9 & 0.01 & 10.1 & 10.2 & 0 \\
Osteoarthritis & 24.2 & 20.3 & 0.09 & 23.2 & 23.1 & 0 \\
Pneumonia & 5.2 & 3.5 & 0.08 & 5 & 5.3 & -0.01 \\
Psoriasis & 1.7 & 1.6 & 0 & 1.6 & 1.7 & -0.01 \\
Rheumatoid arthritis & 10.7 & 5.9 & 0.17 & 10.2 & 10.7 & -0.01 \\
Schizophrenia & 0.1 & 0.3 & -0.04 & 0.1 & 0.2 & -0.02 \\
Ulcerative colitis & 1.1 & 0.7 & 0.04 & 1.1 & 1.1 & 0 \\
Urinary tract infectious   disease & 26.1 & 22.1 & 0.09 & 24.7 & 24.8 & 0 \\
Viral hepatitis C & 0.9 & 0.7 & 0.03 & 0.9 & 1 & -0.01
\end{tabular}}
\end{table}

\begin{table}[p]
\caption{Continued.}
\ContinuedFloat
\resizebox{\columnwidth}{!}{\begin{tabular}{lrrrrrr}
\toprule
& \multicolumn{3}{c}{Before matching} & \multicolumn{3}{c}{After matching} \\
\cmidrule(rl){2-4} \cmidrule(rl){5-7} 
Characteristic & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff & Teriparatide (\%) & Bisphosphonates (\%) & Std. diff \\
\midrule
\textbf{Medical history: Cardiovascular disease} &  &  &  &  &  & \\
Atrial fibrillation & 9.8 & 8.2 & 0.06 & 9.6 & 9.8 & -0.01 \\
Cerebrovascular disease & 2.8 & 2.4 & 0.02 & 2.6 & 2.7 & -0.01 \\
Coronary arteriosclerosis & 8.4 & 7.2 & 0.05 & 8 & 7.9 & 0 \\
Heart disease & 4.4 & 3.7 & 0.03 & 4.1 & 4.1 & 0 \\
Heart failure & 3.7 & 2.8 & 0.05 & 3.5 & 3.7 & -0.01 \\
Ischemic heart disease & 0.3 & 0.2 & 0.03 & 0.3 & 0.2 & 0.02 \\
Pulmonary embolism & 1.4 & 0.8 & 0.05 & 1.3 & 1.4 & -0.01 \\
Venous thrombosis & 0.9 & 0.6 & 0.04 & 0.8 & 0.9 & -0.01 \\
\textbf{Medical history: Neoplasms} &  &  &  &  &  &  \\
Malignant lymphoma & 0.5 & 0.5 & 0 & 0.5 & 0.6 & -0.01 \\
Malignant neoplastic disease & 0.7 & 0.7 & 0 & 0.7 & 0.8 & -0.01 \\
Malignant tumor of colon & 0.1 & 0.1 & 0 & 0.1 & 0.1 & 0 \\
Malignant tumor of lung & 0.1 & 0.1 & 0 & 0.1 & 0.1 & 0 \\
\bottomrule
\end{tabular}}
\end{table}

\end{landscape}


```{r fig-ch7-supp1, cache=TRUE, echo=FALSE, fig.cap="Follow-up distributions of patients treated with teriparatide or bisphosphonates across the four databases.",fig.align='center'}
grid::grid.raster(tiff::readTIFF(here::here("figures/ch7-followup.tiff")))
```

```{r fig-ch7-supp2, cache=TRUE, echo=FALSE, fig.cap="Effect size estimates for the negative controls (true hazard ratio of 1) stratified on baseline hip fracture risk across the four databases.",fig.align='center'}
knitr::include_graphics("figures/ch7-riskStratifiedNcPlot_itt_att_1095_custom.pdf")
```

```{r fig-ch7-supp3, cache=TRUE, echo=FALSE, fig.cap="Effect size estimates for the negative controls (true hazard ratio of 1) stratified on baseline major osteoporotic fracture risk across the four databases.",fig.align='center'}
knitr::include_graphics("figures/ch7-riskStratifiedNcPlot_itt_att_1095_gl.pdf")
```

```{r fig-ch7-supp4, cache=TRUE, echo=FALSE, fig.cap="Hazard ratio estimates of the effect of treatment with teriparatide compared to oral bisphosphonates for the three outcomes of interest (hip fracture, major osteoporotic fracture, and vertebral fracture) stratified on baseline major osteoporotic fracture risk. Estimates are calibrated for unobserved confounding using estimates on 147 negative control outcomes. We plot in red the meta-analytic estimate across the four databases for each of the outcomes of interest. We also plot in yellow the hazard ratio estimated in VERO clinical trial.",fig.align='center'}
grid::grid.raster(tiff::readTIFF(here::here("figures/ch7-plotMetaRiskStratified_itt_att_1095_gl.tiff")))
```

```{r fig-ch7-supp5, cache=TRUE, echo=FALSE, fig.cap="Treatment effect estimates on the absolute scale for the three outcomes of interest stratified on baseline major osteoporotic fracture risk.",fig.align='center'}
grid::grid.raster(tiff::readTIFF(here::here("figures/ch7-plotAbsoluteRiskStratified_itt_att_1095_gl.tiff")))
```

\clearpage\null\pagestyle{empty}
